---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# epizootic

<!-- badges: start -->
[![R-CMD-check](https://github.com/viralemergence/epizootic/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/viralemergence/epizootic/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`epizootic` is an extension to `poems`, a spatially-explicit, process-explicit, pattern-oriented framework for modeling population dynamics. This extension adds functionality for modeling disease dynamics in wildlife. It also adds capability for seasonality and for unique dispersal dynamics for each life cycle stage.

## Installation

You can install the development version of epizootic from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("viralemergence/epizootic")
```

## About R6 classes

`poems` and `epizootic` run on [R6](https://r6.r-lib.org/articles/Introduction.html) classes. R is primarily a *functional* programming language in which the primary units of programming are expressions and functions. Here we use R6 to create an *object-oriented* framework inside of R. R6 classes such as `DiseaseModel` and `SimulationHandler` are used to store model attributes, check them for consistency, pass them to parallel sessions for simulation, and gather results and errors.

## Example

Here is the initial state of an idealized theoretical disease scenario, following a SIR disease model with three life cycle stages:

```{r initial_state}
library(poems)
library(epizootic)
example_region <- Region$new(coordinates = data.frame(x = rep(seq(177.01, 177.05, 0.01), 5),
                             y = rep(seq(-18.01, -18.05, -0.01), each = 5)))
initial_abundance <- c(c(5000, 5000, 0, 1, 0, 0, 0, 0, 0),
                          rep(c(5000, 5000, 0, 0, 0, 0, 0, 0, 0), 24)) |>
      matrix(nrow = 9)
example_region$raster_from_values(initial_abundance[1,]) |>
  raster::plot(main = "Susceptible Juveniles")
example_region$raster_from_values(initial_abundance[4,]) |>
  raster::plot(main = "Infected Juveniles")
```

Here I create a `DiseaseModel` object, which stores inputs for disease simulations and checks them for consistency and completeness.

```{r disease_model}
model_inputs <- DiseaseModel$new(
  time_steps = 10,
    seasons = 2,
    populations = 25,
    stages = 3,
    compartments = 3, # indicates disease compartments
    region = example_region,
    initial_abundance = initial_abundance,
    # Dimensions of carrying_capacity are populations by timesteps
    carrying_capacity = matrix(100000, nrow = 25, ncol = 10),
    # Indicates length of breeding season in days for each population
    breeding_season_length = rep(c(98:102), 5),
    # One mortality value for each stage and compartment
    mortality = c(0.4, 0.2, 0, 0.505, 0.25, 0.105, 0.4, 0.2, 0),
    # Indicates that these are seasonal mortality values
    mortality_unit = 1,
    fecundity = 15,
    fecundity_unit = 1,
    # Indicates that only susceptible and recovered adults reproduce
    fecundity_mask = c(0, 0, 1, 0, 0, 0, 0, 0, 1),
    # Transmission rates from infected individuals, one for each stage
    transmission = c(0.00002, 0.00001, 7.84e-06),
    # Indicates that these are daily transmission rates
    transmission_unit = 0,
    # Indicates that all stages in the first compartment, S, can be infected
    transmission_mask = c(1, 1, 1, 0, 0, 0, 0, 0, 0),
    recovery = c(0.05714286, 0.06, 0.1),
    recovery_unit = 0,
    # Indicates that all stages in the second compartment, I, can recover
    recovery_mask = c(0, 0, 0, 1, 1, 1, 0, 0, 0),
    season_functions = list(sir_model_summer),
    dispersal = list(dispersal_function),
    simulation_order = list(c("transition", "season_functions", "results"),
                            c("dispersal", "results"))
)
```

The core simulation engine of `epizootic` is the function `disease_simulator`, which simulates spatially explicit disease dynamics in populations.
